UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts

### 2019-07-08 ms/cover_letter.tex
more citations to intro, edit caveats a bit, rm arrows from images

\begin{document}
Dear Editor,
\\\\
We are pleased to resubmit our manuscript \emph{UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts} for publication in PLOS Genetics. Using a recently developed dimensionality reduction method called UMAP, this work provides an efficient strategy to identify and visualize population structure in the massive and heterogeneous genomic datasets that are the new frontier of genomics research. In contrast to existing methods, this approach is able to represent both discrete and continuous population structure and the distributions of ancestries, admixture, and phenotypes within the context of such structure.
\\\\
By applying UMAP to large contemporary human genetic datasets (the 1000 Genomes Project, the Health and Retirement Study, and the UK biobank), we reveal a wide range of patterns that had never been observed in these well-studied cohorts. We identify a cluster of Hispanic individuals that is largely restricted to the US Mountain region --- likely descendants of the Hispano population --- and  multiple associations between genetics, geography, and phenotypes in the UK biobank as well as population structure within 1000 Genomes Project populations.
\\\\
We show that UMAP outperforms t-SNE --- the current state-of-the-art --- computationally, qualitatively, and quantitatively, and to such an extent that we expect it to become a standard companion to principal component analysis.
Methods described in our preprint have already been used to study ancestral diversity in the Genome Aggregation Database (gnomAD)\footnote{Karczewski, Konrad J., et al. "Variation across 141,456 human exomes and genomes reveals the spectrum of loss-of-function intolerance across human protein-coding genes." BioRxiv (2019): 531210.} and the National Geographic Genographic Project\footnote{Dai, Chengzhen L., et al. "Population histories of the United States revealed through fine-scale migration and haplotype analysis." BioRxiv (2019): 577411.}, and by 23andMe. Indeed, 23andMe was able to follow up on one of the population structure signals we identified in our preprint by recontacting their participants, and the resulting interpretation of the population structure in the Gujarati is now part of the current manuscript. 
\\\\
Since our initial submission, we have thoroughly edited the manuscript and addressed the comments provided by reviewers. We have added practical discussion on the use of UMAP, caveats to its uses and interpretation , and follow-up on findings suggesting differences in the distribution of FEV1 levels among East Asian individuals in the UK biobank.
\\\\
We believe that this is compelling evidence that UMAP will be useful and of interest to a broad variety of geneticists in population genetics, medical genetics, anthropology, and data science, which makes the manuscript an ideal candidate for PLOS Genetics.
\\
\\
Sincerely,
\\
Alex Diaz-Papkovich and Simon Gravel
\\
on behalf of all authors.

>>>>

\begin{document}
Dear Editor,
\\\\
We are pleased to resubmit our manuscript \emph{UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts} for publication in PLOS Genetics. Since our previous submission we have thoroughly edited the manuscript to address reviewer comments in detail, to provide guidance on usage and interpretation of UMAP, and to provide rigorous follow up on findings.
\\\\
Among the findings are differences between East Asian populations in the distributions of FEV1, a clinically important measurement used in studies of lung function, which have not been identified to date to the best of our knowledge. We have also identified more detailed population structure patterns using the country of birth for individuals in the UK biobank and demonstrated how to use UMAP to identify patterns in ancestry on a global scale.
\\\\
The methods described in our preprint continue to be used to study ancestral diversity, such as in the National Geographic Genographic Project\footnote{Dai, Chengzhen L., et al. "Population histories of the United States revealed through fine-scale migration and haplotype analysis." BioRxiv (2019): 577411.} and by 23andMe\footnote{23andMe. https://blog.23andme.com/ancestry/23andme-tests-new-ancestry-breakdown-in-central-and-south-asia/. ``23andMe Tests New Ancestry Breakdown in Central and South Asia'' (2019-04-03)}, whose interpretation of the population structure in the Gujarati is now part of the current manuscript. 
\\
\\
Sincerely,
\\
Alex Diaz-Papkovich and Simon Gravel
\\
on behalf of all authors.

### 2019-07-08 ms/main.tex
more citations to intro, edit caveats a bit, rm arrows from images

\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}
Questions in medicine, anthropology, and related fields hinge on interpreting the deluge of genomic data provided by modern high-throughput sequencing technologies. Because genomic datasets are high-dimensional, their interpretation requires statistical methods that can comprehensively condense information in a manner that is understandable to researchers and minimizes the amount of data that is sacrificed. Both model-based and model-agnostic approaches to summarize data have played important roles in shaping our understanding of the evolution of our species [e.g., \cite{lawson2012inference, novembre2016recent, spence2018inference}] \sgcomment{Could we put the original PCA paper and structure/admixture, chromopainter?}.

@@

\subsubsection*{Caveats}
The existence of adjustable parameter, including PC cutoff, minimum distance, and number of neighbors, opens the door to tweaking. Figure~\ref{fig:supp_megamontage_pc10_50} shows that disconnected population clusters can appear at different locations depending on parameter choices. Using a minimal number of neighbours (e.g. 5) can result in the formation of disjoint clusters comprised of related family members (figure~\ref{fig:supp_1kgp_families}), and using a low minimum distance can result in clusters becoming more compact, losing visual detail. However, default suggested parameters in UMAP (minimum distance of 0.1 and 15 neighbours) generally performs well across datasets.

@@

Like most non-linear methods, UMAP lacks direct interpretability. It emphasizes local distances over global distances; while points that are very close in UMAP space are likely close in the original data, points that are distant in UMAP space are not necessarily very different in the original data. For these reasons, UMAP coordinates should not be used as GWAS covariates or for quantifying distances between populations. UMAP is sensitive to sample sizes and spends more visual real estate on populations with larger sample sizes. This is useful to identify significant patterns in a cohort, but it makes comparing visualization across cohorts difficult and may appear to exaggerate the genetic variation within the most sampled populations, such as the White British population in the UKBB. We did not assign meaning to wiggles in UMAP figures, which occurred consistently in the UKBB but may be an artefact of the dimensional reduction strategy rather than a meaningful feature of the data. Hand-waving interpretations of pretty plots have a history of getting population geneticists in trouble (as pointed out, e.g., in \cite{novembre2008interpreting}): visualization is not a replacement for statistical testing.

@@

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/montage_fev1_height_chi_eur_permuted.pdf}
    \end{subfigure}
    \caption{Zoomed in section of figure~\ref{fig:umap_ukbb_pc10} focused on individuals with Chinese (CHI), White British (GBR), any other white background, or any other ethnic group (OEG) coloured by ethnicity (left), FEV1 (middle), and age-adjusted height (right). The OEG cluster next to the Chinese cluster appears redder on the middle panel despite similar ages (right panel), suggesting higher FEV1 in that population. A cluster of OEG/other white individuals is bluer relative to the rest of the British or white population, suggesting they may have lower than average FEV1 values relative to the rest of the British or white population.}
    \label{fig:supp_comparison_fev_chi_eur}
\end{figure*}

>>>>

\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}
Questions in medicine, anthropology, and related fields hinge on interpreting the deluge of genomic data provided by modern high-throughput sequencing technologies. Because genomic datasets are high-dimensional, their interpretation requires statistical methods that can comprehensively condense information in a manner that is understandable to researchers and minimizes the amount of data that is sacrificed. Both model-based and model-agnostic approaches to summarize data have played important roles in shaping our understanding of the evolution of our species [e.g., \cite{lawson2012inference, novembre2016recent, spence2018inference, eigen2006, Hellenthal747}].

@@

\subsubsection*{Caveats}
The existence of adjustable parameters, including PC cutoff, minimum distance, and number of neighbours, opens the door to tweaking. Using a minimal number of neighbours (e.g. 5) can result in the formation of disjoint clusters comprised of related family members (figure~\ref{fig:supp_1kgp_families}), and using a low minimum distance can result in clusters becoming more compact, losing visual detail. We used a minimum distance of 0.5 and 15 neighbours; however, default suggested parameters in UMAP (minimum distance of 0.1 and 15 neighbours) generally perform well across datasets.

@@

Like most non-linear methods, UMAP lacks direct interpretability. It emphasizes local distances over global distances; while points that are very close in UMAP space are likely close in the original data, points that are distant in UMAP space are not necessarily very different in the original data. Disconnected clusters may also change their positions relative to other clusters over the course of multiple projections, as in figure~\ref{fig:supp_megamontage_pc10_50}. For these reasons, UMAP coordinates should not be used as GWAS covariates or for quantifying distances between populations. UMAP is sensitive to sample sizes and spends more visual real estate on populations with larger sample sizes. This is useful to identify significant patterns in a cohort, but it makes comparing visualization across cohorts difficult and may appear to exaggerate the genetic variation within the most sampled populations, such as the White British population in the UKBB. We did not assign meaning to wiggles in UMAP figures, which occurred consistently in the UKBB but may be an artefact of the dimensional reduction strategy rather than a meaningful feature of the data. Hand-waving interpretations of pretty plots have a history of getting population geneticists in trouble (as pointed out, e.g., in \cite{novembre2008interpreting}): visualization is not a replacement for statistical testing.

@@

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/montage_fev1_height_chi_eur_permuted.png}
    \end{subfigure}
    \caption{Zoomed in section of figure~\ref{fig:umap_ukbb_pc10} focused on individuals with Chinese (CHI), White British (GBR), any other white background, or any other ethnic group (OEG) coloured by ethnicity (left), FEV1 (middle), and age-adjusted height (right). The OEG cluster next to the Chinese cluster appears redder on the middle panel, suggesting higher levels of FEV1.}
    \label{fig:supp_comparison_fev_chi_eur}
\end{figure*}
